---
title: "Data Wrangling Exercise 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## UCI Data Set

30 subjects performed 6 different activities while wearing a smartphone on their waist.

Subjects were split into 2 groups: 9 were assigned to a 'test' group; 21 to a 'train' group.

The smartphone captured 3 linear acceleration and 3 angular velocity measurements during each of the 6 activities.

Using the recorded measurements, 561 features of each activity were created.

## Data Structure and Organization

9 files of measurements for both the 'test' group and the 'train' group can be found within sub-folders labeled "Inertial Signals".

3 files are contained in each of the 'test' and 'train folders':

1. subject_test/subject_train: a single column of numbers between 1 and 30 identifying the subject, corresponding to the rows of data in the other 2 files 
2. Y_test/Y_train: a single column of numbers between 1 and 6 identifying the activity, corresponding to the rows of data in the other 2 files 
3. X_test/X_train: 561 columns of data representing the 561 features created using the smartphone measurements, corresponding to the rows of data in the other 2 files 

2 further files are provided as reference:

1. activity_labels: a file indicating which activity number corresponds to which activity
2. features: the names of the 561 features created using the smartphoe measurements

## Data Wrangling

### Merge the Test and Train data sets to create one data set

Top begin, the dplyr package was called:

```{r, message = FALSE}
library(dplyr)
```

Column name variables were created to be used when loading the Test and Train data sets:


```{r}
subject_id <- "SubjectID"
activity_number <- "ActivityLabel"
features <- read.table("features.txt")
features_names <- features$V2
```

Next, the 3 Test files were loaded into 3 data frames:

```{r}
subject_test <- read.table("subject_test.txt", col.names = subject_id)
y_test <- read.table("y_test.txt", col.names = activity_number)
x_test <- read.table("X_test.txt", col.names = features_names)
```

These 3 data frames were then combined into a single *test* data frame:
```{r}
test <- bind_cols(subject_test, y_test, x_test)
test[1:6, 1:6]
```

These steps were repeated for the Train files to create a single *train* data frame:
```{r}
subject_train <- read.table("subject_train.txt", col.names = subject_id)
y_train <- read.table("y_train.txt", col.names = activity_number)
x_train <- read.table("X_train.txt", col.names = features_names)
train <- bind_cols(subject_train, y_train, x_train)
train[1:6, 1:6]
```

Finally, the Test and Train data frames were combined into a single data frame called *results*:
```{r}
results <- bind_rows(test, train)
results[1:6, 1:6]
```

### Extract Columns Containing Mean and Standard Deviation
Create a new data frame called *mean_std* by selecting columns from the *results* data frame with names containing *SubjectID*, *ActivityLabel*, *mean*, and *std*

```{r}
mean_std <- select(results, contains("SubjectID"), contains("ActivityLabel"), 
       contains("mean"), contains("std"))
mean_std[1:6, 1:6]
```

### Create Variables ActivityLabel and Activity Name

Load the activity labels file into a data frame to allow it to be merged with the *mean_std* data frame created in the previous step
```{r}
activity_labels <- read.table("activity_labels.txt", 
                              col.names = c("ActivityLabel", "ActivityName"))
```


Merge the *mean_std* data frame with the *activity_labels* data frame by the *ActivityLabel* column
```{r}
mean_std <- merge(mean_std, activity_labels, by = "ActivityLabel")
mean_std[1:6, 1:4]
mean_std[1:6, 86:89]
```

### Create a Final Tidy Data Set with the Average of Each Variable for Each Activity and Each Subject

Create an empty data frame called *activity* to pass calculated values to:
```{r}
activity <- data.frame()
```

Create a loop to select values for each activity and subject, then find mean for each value.
The *ActivityLabel* variable had to be removed for the *mean* function to work (this was added back in later). A new row was added to the *activity* data frame for each activity/subject mean value:
Create an empty data frame called *activity* to pass calculated values to:
```{r}
for(m in 1:6) {
for(n in 1:30) {
   activity <- rbind(activity, summarise_each(filter(mean_std, ActivityLabel == m, SubjectID == n), 
                                    funs(mean), ActivityLabel:fBodyBodyGyroJerkMag.std..))
}
}
activity[1:6, 1:6]
```

The *ActivityLabel* variable was added back into the *activity* data frame in the same manner as used previously:
Create an empty data frame called *activity* to pass calculated values to:
```{r}
activity <- merge(activity, activity_labels, by = "ActivityLabel")
activity[1:6, 84:89]
```

Finally, the *activity* data frame was written to a csv file: 
```{r}
write.csv(activity, file = "wrangling3.csv")
```
